# Youth in the Canadian Labour Market


```{r data-and-packages}
library(censuspumf)
library(tidyverse)
library(gt)

d0 <- censuspumf::records |> 
  censuspumf::encode_labels() 
  
d <- d0 |> 
  mutate(
    is_working_age_lgl = as.integer(AGEGRP) |> between(7, 16),
    is_youth_lgl = as.integer(AGEGRP) |> between(7, 9),
    is_youth_fct = factor(is_youth_lgl * -1 + 2, labels = c("Youth (18-29)", "Not Youth (30-64)")),
    lf_status = LFACT |> str_extract("Employed|Unemployed|Not in the labour force") |> fct_relevel("Employed", "Unemployed", "Not in the labour force"),
    lf_status_neet = case_when(
      lf_status %in% c("Unemployed", "Not in the labour force") & ATTSCH == "Did not attend school" ~ "Not in Employment, Education, or Training (NEET)",
      TRUE ~ as.character(lf_status)
    ) |> 
    fct_relevel("Employed", "Unemployed", "Not in the labour force", "Not in Employment, Education, or Training (NEET)")
  )

dl_button <- function(x, name) {
  x |> 
    downloadthis::download_this(
      output_name = glue::glue("esdc-youth_{name}"), 
      output_extension = ".csv", 
      button_label = "Download data as CSV", 
      button_type = "default", 
      has_icon = TRUE, 
      icon = "fa fa-save"
    )
}
```

## Youth Representation in the Canadian Population

- In this analysis, Youth are defined as individuals aged 18 to 29. We care about this group because they are old enough that most of them have a relationship to the labour market, but young enough that they are still in the process of forming their careers and lives. 
- Youth, as defined, comprise roughly 15% of the Canadian population and 24% of the working-age population.

```{r youth-rep}
t_rep_pop <- d |> 
  filter(!is.na(AGEGRP)) |> 
  group_by(AGEGRP, is_youth_lgl, is_working_age_lgl) |> 
  summarize(n = sum(WEIGHT), .groups = "drop") |> 
  mutate(
    rep_pop = n / sum(n), cumulative_pop = cumsum(rep_pop * is_youth_lgl), 
    rep_working_age = (n * is_working_age_lgl) / sum(n * is_working_age_lgl), 
    cumulative_working_age = cumsum(rep_working_age * is_youth_lgl)
  )

t_rep_pop |> 
  gt() |> 
  fmt_percent(columns = c(rep_pop:cumulative_working_age), decimals = 0)

dl_button(t_rep_pop, "rep")
```

## Youth Distribution Across Provinces

The distribution of youth across provinces is very similar to the distribution of the general population.

```{r youth-distribution}
t_prov_youth <- d |> 
  filter(!is.na(PR), is_youth_lgl) |> 
  group_by(PR) |> 
  summarize(n = sum(WEIGHT), .groups = "drop") |> 
  mutate(`Youth 18-29 (%)` = n / sum(n)) |> 
  rename(`Youth 18-29 (n)` = n)

t_prov_all <- d |> 
  filter(!is.na(PR)) |> 
  group_by(PR) |> 
  summarize(n = sum(WEIGHT), .groups = "drop") |>
  mutate(`All Ages (%)` = n / sum(n)) |> 
  rename(`All Ages (n)` = n)


t_prov_youth |> 
  left_join(t_prov_all, by = "PR") |> 
  gt(rowname_col = "PR") |> 
  fmt_number(columns = c(2, 4), decimals = 0) |> 
  fmt_percent(columns = c(3, 5), decimals = 0)

dl_button(t_prov_youth, "prov")
```

## Youth Demographics

- Compared to other Canadians of working age, higher proportions of youth are:
  - Indigenous (6% vs 4%)
  - Non-immigrants (75% vs 69%)
  - Non-white (41% vs 31%), particularly South Asian (11% vs 7%) and Black (5% vs 3%)

```{r demographics}

t_demos <- d |> 
  pivot_longer(
    cols = c(IMMSTAT, Gender, ABOID, DPGRSUM), 
    names_to = "var", values_to = "value"
  ) |> 
  filter(!is.na(AGEGRP), is_working_age_lgl, !is.na(value)) |>  
  group_by(is_youth_fct, var, value) |>
  summarize(n = sum(WEIGHT), .groups = "drop_last") |> 
  mutate(
    p = n / sum(n)
  )  |> 
  select(-n) |> 
  pivot_wider(names_from = is_youth_fct, values_from = p) 

t_demos |> 
  left_join(
    censuspumf::var_labels
  ) |> 
  group_by(lab) |> 
  select(-var) |> 
  gt(groupname_col = "lab", rowname_col = "value") |> 
  fmt_percent(columns = c(2, 3), decimals = 0)

dl_button(t_demos, "demos")
```


## Youth Education and Labour Force Status

Compared to the working-age population aged 30 or older, youth are:
  - More likely to be unemployed (12% vs 7%)
  - More likely to be attending school (41% vs 5%), especially at the university level (22% vs 2%)
  - Less likely to be "NEET" (15% vs 25%)

```{r lf-status}
t_lfs <- d |> 
  filter(is_working_age_lgl) |> 
  pivot_longer(
    cols = c(ATTSCH, HDGREE, lf_status, lf_status_neet), 
    names_to = "var", values_to = "value"
  ) |> 
  filter(!is.na(value)) |> 
  group_by(is_youth_fct, var, value) |> 
  summarize(n = sum(WEIGHT), .groups = "drop_last") |>
  mutate(
    p = n / sum(n)
  ) |> 
  select(-n) |> 
  pivot_wider(names_from = is_youth_fct, values_from = p) |> 
  left_join(
    censuspumf::var_labels |> 
      filter(var != "ATTSCH") |> 
      bind_rows(
        tibble(
          var = c("lf_status", "lf_status_neet", "ATTSCH"),
          lab = c("Labour Force Status", "Labour Force Status (NEET)", "Education: Ongoing Schooling")
        )
      )
  ) |> 
  ungroup() |> 
  arrange(lab) |> 
  group_by(lab)

t_lfs |> 
  select(-var) |> 
  gt(groupname_col = "lab", rowname_col = "value") |> 
  fmt_percent(columns = c(2, 3), decimals = 0)


dl_button(t_lfs, "lfs")
```

### Intra-youth Comparison of Education and Labour Force Status

```{r intra-youth-lf-status}
t_lfs_iy <- d |> 
  filter(as.integer(AGEGRP) %in% 7:10) |> 
  mutate(
    educ_num = as.integer(HDGREE),
    educ_short = factor(
      case_when(
        educ_num == 1 ~ "No HS", 
        educ_num %in% 2:8 ~ "HS+", 
        educ_num > 8 ~ "Bachelor+"
      ) |> 
      fct_rev()
    )
  ) |> 
  pivot_longer(
    cols = c(ATTSCH, educ_short, HDGREE, lf_status, lf_status_neet), 
    names_to = "var", values_to = "value"
  ) |> 
  filter(!is.na(value)) |> 
  group_by(AGEGRP, var, value) |> 
  summarize(n = sum(WEIGHT), .groups = "drop_last") |>
  mutate(
    p = n / sum(n)
  ) |> 
  select(-n) |> 
  pivot_wider(names_from = AGEGRP, values_from = p, values_fill = 0) |> 
  left_join(
    censuspumf::var_labels |> 
      filter(var != "ATTSCH") |> 
      bind_rows(
        tibble(
          var = c("educ_short", "lf_status", "lf_status_neet", "ATTSCH"),
          lab = c("Education: Condensed", "Labour Force Status", "Labour Force Status (NEET)", "Education: Ongoing Schooling")
        )
      )
  ) |> 
  ungroup() |> 
  arrange(lab) |> 
  group_by(lab)

t_lfs_iy |> 
  select(-var) |> 
  gt(groupname_col = "lab", rowname_col = "value") |> 
  fmt_percent(columns = c(2, 3, 4, 5), decimals = 0)


dl_button(t_lfs_iy, "lfs-iy")
```

### Employment Characteristics

```{r job-features}
t_work <- d |> 
  filter(is_working_age_lgl) |> 
  pivot_longer(
    cols = c(FPTWK, PWDUR, MODE, POWST), 
    names_to = "var", values_to = "value"
  ) |> 
  filter(!is.na(value)) |> 
  group_by(is_youth_fct, var, value) |> 
  summarize(n = sum(WEIGHT), .groups = "drop_last") |>
  mutate(
    p = n / sum(n)
  ) |> 
  select(-n) |> 
  pivot_wider(names_from = is_youth_fct, values_from = p) |>
  mutate(var = str_to_upper(var)) |> 
  left_join(
    censuspumf::var_labels, by = "var" 
  ) |> 
  ungroup() |> 
  arrange(lab) |> 
  group_by(lab)

d |> 
  select(where(~is.factor(.x) & any(str_detect(levels(.x), "\\$"))))

t_work |> 
  select(-var) |> 
  gt() |> 
  fmt_percent(columns = c(2, 3), decimals = 0)

dl_button(t_work, "work")
```


### Income

Average total income for working-age youth is roughly half of that for the 30+ working-age population. 

```{r income}
t_income <- d |> 
  filter(is_working_age_lgl) |> 
  pivot_longer(
    cols = c(EmpIn, Invst, GTRfs, TotInc), 
    names_to = "var", values_to = "value"
  ) |> 
  filter(!is.na(value)) |> 
  group_by(is_youth_fct, var) |> 
  summarize(n = sum(WEIGHT), mean = weighted.mean(value, WEIGHT), .groups = "drop") |> 
  mutate(
    var = str_to_upper(var)
  ) |> 
  left_join(var_labels) |> 
  select(-var) |> 
  group_by(lab)

t_income |> 
  gt(groupname_col = "lab", rowname_col = "is_youth_fct") |> 
  fmt_number(columns = n, decimals = 0) |> 
  fmt_currency(columns = "mean", decimals = 0)

dl_button(t_income, "income")
```

### Key Finding: The Canadian Labour Force is Growing More Educated

We observe a smaller proportion of youth who have attained a university degree (24%) compared to non-youth of working age (31%), but the much higher proportion of working-age youth in university suggests that this gap may invert in the future. 

### Key Finding: The NEET Youth Population is Too Large to Ignore

Of the working-age youth who are not employed, most are in school. 
A quarter of the 30+ working-age population is NEET, as compared to 15% of working-age youth.
In absolute terms, however, the number of NEET youth is still substantial (809,115). 

```{r neetness}
t_neet <- d |> 
  filter(is_youth_lgl, !is.na(lf_status_neet)) |> 
  group_by(lf_status_neet) |> 
  summarize(
    n = sum(WEIGHT)
  ) 
  
t_neet |> 
  gt() |> 
  fmt_number(columns = "n", decimals = 0)

dl_button(t_neet, "neet")
```

## Poverty and Hardship

According to all of the low-income measures provided in the Census PUMF, working-age youth are more likely than their older counterparts to be live in income poverty. According to the MBM poverty line (Canada's current official measure), 12% of working-age youth are in poverty, compared to 8% of the 30+ working-age population. In spite of this, working-age youth are no more likely to be in core housing need than their elders. Across both groups, the rate of core housing need is 7%. 

```{r poverty-etc}
t_pov <- d |> 
  filter(is_working_age_lgl) |> 
  pivot_longer(
    cols = c(matches("LICO|LIM|LoMBM"), HCORENEED_IND), 
    names_to = "var", values_to = "value"
  ) |> 
  mutate(
    var = str_to_upper(var)
  ) |> 
  filter(!is.na(value)) |>
  group_by(is_youth_fct, var, value) |>
  summarize(
    n = sum(WEIGHT), .groups = "drop_last"
  ) |>
  mutate(
    p = n / sum(n)
  ) |> 
  select(-n) |> 
  pivot_wider(names_from = is_youth_fct, values_from = p) |> 
  ungroup() |> 
  left_join(
    censuspumf::var_labels
  ) |> 
  group_by(lab)

t_pov |> 
  select(-var) |> 
  gt(groupname_col = "lab", rowname_col = "value") |> 
  fmt_percent(columns = c(2, 3), decimals = 0)

dl_button(t_pov, "poverty")
```

